\vspace{-5mm}
{\Large \sffamily Acknowledgements}
\vspace{3mm}

The course of my last four years until the very last moment of the writing of this dissertation has been nothing but a thrilling ride. 
A ride that felt at times as smooth as on a highway, sometimes like a steep mountain climb, other times as if I was on vast plain with no particular direction, and sometimes like exploring in a cave that needed some explosions here and there. Wherever the road went though, if there is one thing that I learned, it is that the way gives its rewards as long as I keep moving on. It is a relief looking back at it right now and I feel gratitude for many people that accompanied me on this journey. 

First of all, I would like to give thanks to my supervisors that assisted me along the way. Thank you Mireia for always believing in my capabilities and ideas. Your positive attitude and unconditional support has empowered me greatly in arriving to this point. Toni, I feel extremely lucky to have had the opportunity to work closely with you. I have learned so much in this period not just in terms of technical and methodical knowledge but also in terms of collaboration skills.

It was a priviledge to work besides the members of the TALN tribe this past four years. People that are so talented, hard-working and some also exceptionally good at partying. A round of special thanks goes to Francesco, Luis, Juan and Mónica for their key advices and companionship, and to Miguel, Roberto, Simon and Alicia for being the game-changers in helping me keep pace and help me set a direction when I needed it. 

My study would not have been possible without the financial support of DTIC and the workers that maintain it. Special thanks to Aurelio Ruiz for his support and also the management of the DTIC-Maria de Maetzu program which contributed an important share to my research. 

A great deal of motivation that help me realize my work has to be attributed to my colleagues in Col·lectivaT: Pelin, Federica, Güneş and Baybars. I feel very lucky and hopeful for the future collaborating with you.

I keep a deep gratitude for the city of Barcelona and all the people in it that made me feel home here during the course of my PhD: Carla, Bora, Kerim, Berkay, Batu, Nico, Cam, Lauri, Lisa, Paco, Kiket and Mona. 

A big, heartful thanks to my family for always believing in the path that I take and offering their support in every means possible. Thank you Abi, for showing me how cool science can be; Baba, for your pride in my actions and ideas; and Anne for your unconditional love and relentless call to joy.

And finally, my family here, Carmen. I owe much to the faith you have in me in realizing my dreams. Thank you for making it getting better all the time, even this dissertation. 


